---
title: "Political Rents Under A Changing Electoral System"
author: |
  | Bas Machielsen
  | a.h.machielsen@uu.nl
institute: "Utrecht University"
email: "a.h.machielsen@uu.nl"
date: \today
output: 
  beamer_presentation:
    includes:
      in_header: preamble.tex
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(stargazer)
library(tidyverse)
library(ivreg)
library(knitr)

source("helper.R")
```

# Motivation

## Motivation

- There is strong empirical evidence that politicians use public office to serve private interests. They can do so in various ways:
  - Monetary (e.g. Eggers & Hainmuller, 2009; Fisman et al., 2014)
  - Nepotism (Dal Bo et al., 2009; Fafchamps & Labonne, 2017, Folke et al., 2017)
  - Ideology (Mian et al., 2010)

- The literature that investigates under which circumstances politicians can accrue political rents. Factors that matter:
  - High-corruption environment (Fisman et al., 2014)
  - Electoral competition and the media (Svaleryd & Vlachos, 2009)
  - Opaque public procurement (Baltrunaite, 2020)
  
- Contemporary literature teaches that features of the political system influence the degree to which politicians can engage in rent-seeking activity. 

## This Study

- Most studies are static in nature and do not focus on the institutional determinants of political rents. 
  
- This study focuses on a _dynamic_ environment in which the political system changes: the case of the Netherlands (1860-1917)

- Many aspects of the political system changed: 
  - The rise of the career politician (Career Paths)
  - Political party formation and party politics
  - Suffrage extensions and universal suffrage

- Setting allows me to investigate the influence of changing institutions on the magnitude of political rents available to politicians.

## The case of the Netherlands

- 19th century political changes comparable to many other (European) countries 
  - Repeated suffrage extensions (1887, 1896) culminating in universal suffrage
  - From political factions to explicit political parties (Protestants, Catholics, Liberals)
  
- Bicameral system: a Lower House and an Upper House
  - Politics concentrated in Lower House: from 75 to 100 seats
  - Modest formal salary (2500 guilders)

- Elections on the basis of a district system at the time $\rightarrow$ Many (close) elections
  
- I employ a regression discontinuity methodology based on close elections to estimate political rents using the method of Cattaneo et al. (2019)
  - Detailed data on characteristics of politicians to examine what allows politicians to accrue rents

## Results - Overview

- I first replicate the methodology from the contemporary literature to show the existence of politcal rents in the 19th century Netherlands:
  
```{r out.width="80%", fig.align='center'}
knitr::include_graphics("../Tables/RDD_Plot.pdf")
```

- And then set out out to find the influence of career paths, party politics & organization and suffrage extensions on political rents. 

# Data & Methodology

## Sources

- Elections: _Repositorium Tweede Kamerverkiezingen_ (Repository Lower House Elections) 
  - Contains detailed data on every election that took place in the district system (1848-1917)
  - Includes lists of candidates for each election, and amount of votes
  - On this basis, I calculate the Margin and find candidates in close elections
  
- Probate inventories: _Memories van Successie_
  - Archival source available from 1877-1921
  - Measure of political rents: wealth at death
  - Finding rate: about 70% of inventories
  - Main reason for absence: probate inventory registered in other place than place of death.

## Other Sources

- _HDNG Database_, containing information about Dutch Municipalities, including demographics, religious composition, taxes levied, professional composition at various points in time
  - Used to extract birthplace and district characteristics
  
- _Newspaper recommendations_:
  - Newspaper recommendations of each candidate, turnout, no. of times participated, year of election, extracted from the _Repositorium_
  - Used to extract several other election- and candidate characteristics
  
- _Politiek Documentatie Centrum_:
  - Used to find demographic characteristics for politicians
  - Age at election, lifespan, year of death
  - Nonpoliticians: from _Delpher_ and genealogy websites

## Close Elections

- I take into account multi-candidate elections (cf. Lee, 2008)

- Somewhat more complicated definition of the running variable  _Margin_. For candidate $i$ in election $e$:
  
$$
\text{Margin}_{i,e} = \frac{\text{Votes}_{i,e} - \text{Votes}_{j,e}}{\text{Total Votes}_e}
$$
where $j$ is the _Marginal Loser_ if $i$ is a _Winning Contender_, and $j$ is the _Marginal Winner_ if $i$ is a _Losing Contender_. 

- I then regard an individual as having taken part in a close election if $|\text{Margin}_{i,e}| < 0.2$. Using this criterion, about 600 close elections (out of approx. 2400)

## Method

- I use the methodology by Cattaneo et al. (2019), who implement a local linear MSE-optimal RD Estimator

- This estimator optimally trades off bias and variance when picking the bandwidth
  - Consistent with recommendations of Imbens and Gelman (2018)
  - Bias-corrected (Robust) confidence intervals for inference
  
- Specification:

  $$
     g(W_{i}) = \alpha + \delta \cdot 1_{\text{Margin}_i > 0}+ \eta \cdot \text{Margin}_i + X_i \beta + \epsilon_i
  $$
- I use two variants of $g$: $g(.) = \log (.)$ and $g(.) = \text{ihs}(.)$ 
  
# Analysis

## Covariate Balance

- The identifying assumption of the design implies that the control group (non-politicians) should be similar to the politicians at the margin: this gives evidence that the potential outcomes are not discontinuous at the cut-off point. 


```{r}
tab1
```


## Covariate Balance - Cont'd

```{r}
tab2
```

## Geographic Variation

- The elections also seem to be spread out geographically:

```{r}
knitr::include_graphics("../Data/shapefiles/close_elections_geographic.pdf")
```

## Main Results

- The results show a significant and positive effect, irrespective of the inclusion of several covariates. 

```{r, out.width="60%"}
tab3
```

# Mechanisms: Career Paths

## Career Paths

- Did politicians accrue rents by means of their network? Were they able to capure privileged, exclusive positions afterwards? 
  - Most of the politicians staying in politics stayed in the Lower House or became _burgemeesters_

```{r}
tab5
```
## Career Paths

- Politicians who stayed in the Lower House for a long time were also able to collect more rents
  - I analyze the rents acquired by politicians with a longer period of stay in the Lower House compared to those with a shorter period of stay (_Tenure_)

```{=latex}
\begin{table}[!h]

\centering
\fontsize{7}{9}\selectfont
\begin{threeparttable}
\begin{tabular}[t]{lllll}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Log(Wealth)} & \multicolumn{2}{c}{Ihs(Wealth)} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
\multicolumn{1}{c}{ } & \multicolumn{1}{c}{Tenure > 20} & \multicolumn{1}{c}{Tenure < 5} & \multicolumn{1}{c}{Tenure > 20} & \multicolumn{1}{c}{Tenure < 5} \\
\cmidrule(l{3pt}r{3pt}){2-2} \cmidrule(l{3pt}r{3pt}){3-3} \cmidrule(l{3pt}r{3pt}){4-4} \cmidrule(l{3pt}r{3pt}){5-5}
  & (1) & (2) & (3) & (4)\\
\midrule
Coefficient & 3.087 & 1.897 & 3.528 & 2.566\\
SE (BC) & (1.541)** & (0.954)** & (2.083)* & (2.069)\\
SE (Rob.) & (1.800)* & (1.169) & (2.314)* & (2.401)\\
N (Politicians) & 44 & 72 & 50 & 85\\
N (Non-Politicians) & 241 & 241 & 245 & 245\\
Bandwidth & Optimal & Optimal & Optimal & Optimal\\
\bottomrule
\end{tabular}
\end{threeparttable}
\end{table}

```

# Mechanisms: Party Organization

## Party Organization - Differences between parties

- I focus on the difference _between_ parties. 
- Very large effect for Protestant politicians, intermediate for liberals and invisible for Catholic politicians. 

```{r}
knitr::include_graphics("../Tables/rdplot_per_party.pdf")
```

## Party Organization - Within Parties

- I exploit the establishment of political parties to find out whether politicians with the same political allegiance have been able to accumulate more or less rents after establishment (and joining of) a political party.
  - I can do this using newspaper recommendations to identify party alignment

```{r}
tab4
```

# Mechanisms: Suffrage Extensions

## Electoral Competition over Time

- Do political rents change significantly following suffrage extensions? 
  - The graph shows CI's based on bootstrapped standard errors. 

```{r out.width="80%", fig.align='center'}
knitr::include_graphics("../Tables/electoral_competition.pdf")
```

## Electoral Competition - Turnout

- At the level of the district, does more turnout mean more monitoring, and lower rents? 
  - I estimate the difference in rents between upper-quantile observations (in terms of relative turnout) and lower-quantiles. 

```{r out.width="80%", fig.align='center'}
knitr::include_graphics("../Tables/turnout_competition.pdf")
```

# Conclusion

## Conclusion

- Politicians who just won elections are substantially wealthier at the end of their life than their contenders who just lost, even though pre-election, they were equal in many aspects. 
  
- Rewards obtained through staying longer in politics
  - Access to other political functions with discretion
  - Possibility of rent-seeking, superior information

- Political rents mainly concentrated among Protestant and Liberal politicians
  - Large contrast with Catholics, possibly because of low value of rents for Catholic politicians

- These politicians are not constrained (enough) by their party 
  - Consistent with voting behavior serving as a signal, or as catering to interest groups 
  - Still, political parties limit voting freedom and thus the possibility to accrue rents

- Electoral discipline (turnout) at individual elections plays a role

# Robustness Checks

## Sensitivity of Estimates to Bandwidth

- Standard estimates (with covariates) appear not to be sensitive to the particular bandwidth choice:

```{r}
knitr::include_graphics("../Tables/bandwidth_test.pdf")
```

## Density Plot

```{r}
knitr::include_graphics("../Tables/Density_Plot_MarginWealth.pdf")
```

## Wealth Densities per Party and Status

```{r}
knitr::include_graphics("../Tables/histogram_wealth_party.pdf")
```

## Flexible Optimal Bandwidth on Both Sides

```{=latex}
\begin{table}[!h]
\centering
\fontsize{7}{9}\selectfont
\begin{threeparttable}
\begin{tabular}[t]{lllll}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Log(Wealth)} & \multicolumn{2}{c}{Ihs(Wealth)} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
  & (1) & (2) & (3) & (4)\\
\midrule
\addlinespace[0.3em]
\multicolumn{5}{l}{\textbf{Panel A: Baseline Estimates}}\\
\hspace{1em}Coefficient & 1.951 & 2.004 & 2.989 & 2.217\\
\hspace{1em}SE (BC) & (0.750)*** & (0.529)*** & (1.272)*** & (0.946)***\\
\hspace{1em}SE (Rob.) & (0.906)** & (0.686)*** & (1.571)** & (1.184)**\\
\hspace{1em}Mean DV Politicians (1\%) & 11.846 & 11.846 & 11.888 & \vphantom{1} 11.888\\
\hspace{1em}Mean DV Non-Politicians (1\%) & 10.134 & 10.134 & 9.504 & \vphantom{1} 9.504\\
\hspace{1em}N (Politicians) & 323 & 323 & 348 & 348\\
\hspace{1em}N (Non-Politicians) & 258 & 258 & 263 & 263\\
\hspace{1em}Bandwidth & Optimal & 2 x Optimal & Optimal & 2 x \vphantom{1} Optimal\\
\addlinespace[0.3em]
\multicolumn{5}{l}{\textbf{Panel B: Estimates With Selected Covariates}}\\
\hspace{1em}Coefficient & 2.055 & 1.715 & 2.728 & 1.399\\
\hspace{1em}SE (BC) & (0.701)*** & (0.477)*** & (0.804)*** & (0.674)***\\
\hspace{1em}SE (Rob.) & (0.931)** & (0.697)*** & (1.308)** & (1.077)*\\
\hspace{1em}Mean DV Politicians (1\%) & 11.846 & 11.846 & 11.888 & 11.888\\
\hspace{1em}Mean DV Non-Politicians (1\%) & 10.134 & 10.134 & 9.504 & 9.504\\
\hspace{1em}N (Politicians) & 254 & 254 & 275 & 275\\
\hspace{1em}N (Non-Politicians) & 249 & 249 & 253 & 253\\
\hspace{1em}Bandwidth & Optimal & 2 x Optimal & Optimal & 2 x Optimal\\
\bottomrule
\end{tabular}
\end{threeparttable}
\end{table}

```

## Metropolitan vs. Rural Areas

- Political Rents for politicians who have been born inside and outside a radius of $x$ km of the Hague (governmental capital)
  - Politicians born within the Randstad (radius of about 60/70 km within the Hague) garner much more rents than politicians born outside. 

```{r}
knitr::include_graphics("../Tables/randstad_distance.pdf")
```

## Flexible Optimal Bandwidth - Career Paths

- I estimate the rents for long-serving (tenure > 20) and short-serving (tenure < 5) politicians in the Lower House. 
  - Politicians who stay longer are able to obtain more rents than politicians who stay shorter.

```{=latex}
\begin{table}[!h]
\centering
\fontsize{7}{9}\selectfont
\begin{threeparttable}
\begin{tabular}[t]{lllll}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Log(Wealth)} & \multicolumn{2}{c}{Ihs(Wealth)} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
\multicolumn{1}{c}{ } & \multicolumn{1}{c}{Tenure > 20} & \multicolumn{1}{c}{Tenure < 5} & \multicolumn{1}{c}{Tenure > 20} & \multicolumn{1}{c}{Tenure < 5} \\
\cmidrule(l{3pt}r{3pt}){2-2} \cmidrule(l{3pt}r{3pt}){3-3} \cmidrule(l{3pt}r{3pt}){4-4} \cmidrule(l{3pt}r{3pt}){5-5}
  & (1) & (2) & (3) & (4)\\
\midrule
Coefficient & 3.533 & 1.872 & 3.579 & 2.442\\
SE (BC) & (1.401)** & (0.890)** & (1.837)* & (1.773)\\
SE (Rob.) & (1.618)** & (1.100)* & (1.988)** & (2.075)\\
N (Politicians) & 44 & 72 & 50 & 85\\
N (Non-Politicians) & 241 & 241 & 245 & 245\\
Bandwidth & Optimal & Optimal & Optimal & Optimal\\
\bottomrule
\end{tabular}
\end{threeparttable}
\end{table}

```

## Flexible Optimal Bandwidth - Party Organization

- These estimates use the optimal bandwidth estimated at each side of the cut-off point
  - The estimates in (4) are again significant at the 10% level

```{=latex}
\begin{table}[!h]
\centering
\fontsize{6}{8}\selectfont
\begin{threeparttable}
\begin{tabular}[t]{lllrllr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{No Covariates} & \multicolumn{3}{c}{With Covariates} \\
\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-7}
\multicolumn{1}{c}{ } & \multicolumn{1}{c}{After} & \multicolumn{1}{c}{Before} & \multicolumn{1}{c}{Diff. (p-value)} & \multicolumn{1}{c}{After} & \multicolumn{1}{c}{Before} & \multicolumn{1}{c}{Diff. (p-value)} \\
\cmidrule(l{3pt}r{3pt}){2-2} \cmidrule(l{3pt}r{3pt}){3-3} \cmidrule(l{3pt}r{3pt}){4-4} \cmidrule(l{3pt}r{3pt}){5-5} \cmidrule(l{3pt}r{3pt}){6-6} \cmidrule(l{3pt}r{3pt}){7-7}
  & (1) & (2) &    & (3) & (4) &    \\
\midrule
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{Panel A: All control observations}}\\
\hspace{1em}Coefficient & 1.977 & 2.178 & -0.143 & 2.29 & 3.555 & -1.265\\
\hspace{1em}SE (BC) & (0.751)*** & (1.006)** & 0.462 & (0.890)*** & (1.071)*** & 0.204\\
\hspace{1em}SE (Rob.) & (0.906)** & (1.214)* &  & (1.078)** & (1.296)*** & \\
\hspace{1em}N Treated & 202 & 121 &  & 202 & 121 & \\
\hspace{1em}N Control & 258 & 258 &  & 258 & 258 & \\
\hspace{1em}Covariates & No & No &  & Yes & Yes \vphantom{1} & \\
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{Panel B: Contemporaneous control observations}}\\
\hspace{1em}Coefficient & 1.594 & 3.098 & -1.504 & 1.939 & 4.876 & -2.937\\
\hspace{1em}SE (BC) & (0.997) & (1.608)* & 0.237 & (0.983)* & (1.767)*** & 0.081\\
\hspace{1em}SE (Rob.) & (1.213) & (1.941) &  & (1.202) & (2.074)** & \\
\hspace{1em}N Treated & 202 & 92 &  & 202 & 92 & \\
\hspace{1em}N Control & 151 & 107 &  & 151 & 107 & \\
\hspace{1em}Covariates & No & No &  & Yes & Yes & \\
\bottomrule
\end{tabular}
\end{threeparttable}
\end{table}

```
