---
title: "Political Rents Under A Changing Electoral System"
author: |
  | Bas Machielsen
  | a.h.machielsen@uu.nl
institute: "Utrecht University"
email: "a.h.machielsen@uu.nl"
date: \today
output: 
  beamer_presentation:
    includes:
      in_header: preamble.tex
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(stargazer)
library(tidyverse)
library(ivreg)
library(knitr)

source("helper.R")
```

# Motivation

## Motivation

- Several studies have shown that politicians benefit from their position. These political rents can be accumulated in a multitude of ways:
  - Monetary (e.g. Eggers & Hainmuller, 2009; Fisman et al., 2014)
  - Nepotism (Dal Bo et al., 2009; Fafchamps & Labonne, 2017)
  - Ideology (Mian et al., 2010)

- Most studies are static in nature and do not focus on the institutional determinants of political rents. 
  - Political rents are likely an equilibrium outcome (e.g. Barro, 1973; Trebbi, 2020)

- Historically, political rents could have affected the democratization process.  
  - How constrained were politicians when implementing universal suffrage step-by-step? (Przeworski, 2009)
  - Political rents have implications for political selection (Besley, 2005) and politicians' voting behavior

## This Study

- I provide a dynamic perspective of political rents by focusing on the case of the Netherlands (1860-1917)

- The Netherlands employed a district system at the time $\rightarrow$ Many (close) elections
  - I employ a regression discontinuity methodology to estimate political rents based on Cattaneo et al. (2019)
  - Detailed data on characteristics of politicians to examine what allows politicians to accrue rents

- Setting allows me to provide insights in the effect on political rents of:
  - Suffrage extensions and the monitoring of politicians
  - The roles of political parties and geography 
  - What career paths politicians followed to pursue rents

## Results - Overview

- At the margin, virtually all covariates are balanced among politicians and non-politicians, making non-politicians a convincing control group

- There is a large and significant positive effect of being elected into politics on personal end-of-life wealth
  - _A priori_ not obvious a politicians would accumulate surplus wealth relative to e.g. businessman, manufacturer, lawyer, etc. 

- The effect seems to be most convincingly present of _Protestant_ politicians, and to a lesser degree _Liberals_
  - These are the same politicians whose parties did not enforce strong voting discipline

- Large discrepancies in estimates of political rents according to regional origin of politicians. 

- Limited evidence of an influence of monitoring mechanisms and suffrage extensions further constraining politicians' rent-seeking activity. 

# Data & Methodology

## Close Elections

- _Repositorium Tweede Kamerverkiezingen_ (Repository Lower House Elections) contains detailed data on every election that took place in the district system (1848-1917)
  - Includes lists of candidates for each election, and amount of votes
  - I take into account multi-candidate elections (cf. Lee, 2008)
  - Somewhat more complicated definition of the running variable  _Margin_. For candidate $i$ in election $e$:
  
  $$
  \text{Margin}_{i,e} = \frac{\text{Votes}_{i,e} - \text{Votes}_{j,e}}{\text{Total Votes}_e}
  $$
where $j$ is the _Marginal Loser_ if $i$ is a _Winning Contender_, and $j$ is the _Marginal Winner_ if $i$ is a _Losing Contender_. 

- I then regard an individual as having taken part in a close election if $|\text{Margin}_{i,e}| < 0.2$. Using this criterion, about 600 close elections (out of approx. 2400)

## Finding Candidates in Archives

- **Politicians**: I use a proprietary dataset from the \textit{Politiek Documentatiecentrum} (PDC)

- The database contains demographic and other variables related to a politicians' life, including their birth and death date and place, as well as information about their political allegiance and their professional career. 

- Using this information, I can find politicians' probate inventories. 

- **Non-Politicians**: For non-politicians, I make use of the search engines of provincial archives, the historical newspaper search engine _Delpher_ and various genealogy websites to find the dates and places of decease. 

## Measure Political Rents

- End of Life Wealth from the _Memories van Successies_ archives
  - Probate inventories assembled with the purpose of levying inheritance taxation
  - Containing a detailed description of the deceased's assets and liabilities, including an appraisal
  - In practice, every individual deceased in the Netherlands (very few exemptions)
  - I use Net Wealth and deflated it to 1900 guilders. 
  
- Freely accessible in the provincial archives from 1877-1927. Outside of this period without appraisal of assets and liabilities (<1877) or not publicly accessible (1927-). 

- Finding rate: about 70% of inventories
  - Main reason for absence: probate inventory registered in other place than place of death.

## Other Data

- Birthplace and district characteristics
  - HDNG Database, containing information about Dutch Municipalities, including demographics, religious composition, taxes levied, professional composition at various points in time
  
- Election characteristics:
  - Newspaper recommendations of each candidate, turnout, no. of times participated, year of election, on the basis of the _Repositorium_
  
- Demographic characteristics:
  - Age at election, lifespan, year of death from the PDC and genealogy websites
  
## Method

- I use the methodology by Cattaneo et al. (2019), who implement a local linear MSE-optimal RD Estimator

- This estimator optimally trades off bias and variance when picking the bandwidth
  - Consistent with recommendations of Imbens and Gelman (2018)
  - Bias-corrected (Robust) confidence intervals for inference
  
- Specification:

  $$
     g(W_{i}) = \alpha + \delta \cdot 1_{\text{Margin}_i > 0}+ \eta \cdot \text{Margin}_i + X_i \beta + \epsilon_i
  $$
- I use two variants of $g$: $g(.) = \log (.)$ and $g(.) = \text{ihs}(.)$ 
  
# Analysis

## Covariate Balance

- The identifying assumption of the design implies that the control group (non-politicians) should be similar to the politicians at the margin: this gives evidence that the potential outcomes are not discontinuous at the cut-off point. 


```{r}
tab1
```


## Covariate Balance 

```{r}
tab2
```

## Geographic Variation

- The elections also seem to be spread out geographically:

```{r}
knitr::include_graphics("../Data/shapefiles/close_elections_geographic.pdf")
```

## Main Results

- The results show a significant and positive effect, irrespective of the inclusion of several covariates. 

```{r, out.width="60%"}
tab3
```

## Main Results (Graphically)

```{r}
knitr::include_graphics("../Tables/RDD_Plot.pdf")
```

# Mechanisms

## Party Organization - Differences between parties

- First, I focus on the difference between parties. 
- Very large effect for Protestant politicians, intermediate for liberals and invisible for Catholic politicians. 

```{r}
knitr::include_graphics("../Tables/rdplot_per_party.pdf")
```

## Party Organization - Within Parties

- I exploit the establishment of political parties to find out whether politicians with the same political allegiance have been able to accumulate more or less rents after establishment (and joining of) a political party.

```{r}
tab4
```


## Career Paths

- Did politicians accrue rents by means of their network? Were they able to capure privileged, exclusive positions afterwards? 

```{r}
tab5
```


## Electoral Competition over Time

- Do political rents change significantly following suffrage extensions? 
  - The graph shows CI's based on bootstrapped standard errors. 

```{r out.width="80%", fig.align='center'}
knitr::include_graphics("../Tables/electoral_competition.pdf")
```

## Electoral Competition - Turnout

- At the level of the district, does more turnout mean more monitoring, and lower rents? 
  - I estimate the difference in rents between upper-quantile observations (in terms of relative turnout) and lower-quantiles. 

```{r out.width="80%", fig.align='center'}
knitr::include_graphics("../Tables/turnout_competition.pdf")
```

## Metropolitan vs. Rural Areas

- Political Rents for politicians who have been born inside and outside a radius of $x$ km of the Hague (governmental capital)
  - Politicians born within the Randstad (radius of about 60/70 km within the Hague) garner much more rents than politicians born outside. 

```{r}
knitr::include_graphics("../Tables/randstad_distance.pdf")
```

# Conclusion

## Conclusion

- Protestant politicians
  - A lot of voting freedom
  - Although parties limit voting freedom and thus rents
  
- Rewards: staying longer in politics - access to other political functions with discretion, possibility of rent-seeking

- Primarily concentrated for politicians within a certain region
  - Especially large contrast with Catholic politicians, not from this region

- Electoral discipline and monitoring plays very small role


